---
title: "Prediction of Martensite Start Temperature in Steels Using Cross Validation"
author: "Mounika Chevva (Advisor: Dr. Samantha Seals)"
date: " 10/06/2024"
format:
  html:
    code-fold: true
course: IDC6490 Capstone in Data Science
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

#### Week 7 (full (polished) description and exploratory data analysis / visualization of the dataset chosen)

```{r}
library(readr)

Martensite <- read.csv("/Users/LaptopUser/Downloads/dataset.csv/Martensite Start-Temp.csv")

head(Martensite)
str(Martensite)


```

####  Dataset Description

The martensite start temperature (Ms) is a critical parameter when designing high-performance
steels and their heat treatments.

Martensite start temperature (Ms) is our target variable.

This table categorizes the variables based on their type and description to our study.


```{r}
library(knitr)

var_desc <- data.frame(
  Var = c("C", "Mn", "Si", "Cr", "Ni", "Mo", "V", "Co", "Al", 
               "W", "Cu", "Nb", "Ti", "B", "N", "Ms"),
  Description = c(
    "Carbon ",
    "Manganese ",
    "Silicon ",
    "Chromium ",
    "Nickel ",
    "Molybdenum ",
    "Vanadium",
    "Cobalt ",
    "Aluminum",
    "Tungsten ",
    "Copper ",
    "Niobium ",
    "Titanium ",
    "Boron ",
    "Nitrogen ",
    "Martensite Start Temperature Target Variable(°C)"
  ),
  Type = c(
    "Numeric","Numeric","Numeric", "Numeric", "Numeric", "Numeric", "Numeric", "Numeric",
   "Numeric", "Numeric", "Numeric", "Numeric", "Numeric", "Numeric", "Numeric", "Numeric"
  ),
  stringsAsFactors = FALSE  
)

kable(var_desc)


```


#### Exploratory Data Analysis (EDA)

Exploratory Data Analysis (EDA) on the Martensite Start Temperature (Ms) dataset involves understanding the distribution, relationships, and patterns in the data.

```{r}
summary(Martensite)
sum(is.na(Martensite))

```

The output [1] 0 from sum(is.na(Martensite)) indicates that there are no missing values (NAs) in the Martensite dataset. This means that all entries in the dataset have complete data without any gaps.

### Data Visualization

#### Histogram for Martensite Start Temperature 

```{r}
library(ggplot2)
ggplot(Martensite, aes(x = Ms)) +
  geom_histogram(binwidth = 10, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Martensite Start Temperature (Ms)", x = "Martensite Start Temperature", y = "Frequency")
```
The histogram describes  the distribution of Martensite Start Temperature (Ms) values, ranging from 200 to 800°C. It shows a bimodal distribution with peaks around 400-450°C and 600-650°C, indicating two predominant temperature ranges. 

#### Correlation matrix 

The correlation matrix is a powerful tool used to visualize and quantify the relationships between key variables

```{r}
library(ggplot2)
library(reshape2)
library(corrplot)

correlation_matrix <- cor(Martensite, use = "complete.obs")

print(correlation_matrix)

corrplot(correlation_matrix, method = "circle")
```

#### Insights from correlation matrix 

The matrix uses color intensity to indicate the strength of the relationships between variables.

Correlation coefficients displayed in the matrix range from -1 to 1, where values near ±1 indicate strong positive or negative correlations, and values near 0 indicate weak or no linear relationship.

The correlations help identify significant predictors of Martensite:

C (Carbon) shows a strong positive correlation with Mn (Manganese), suggesting that as carbon content increases, manganese also tends to increase.

Ms (Martensite Start Temperature) has several notable negative correlations with elements like V (Vanadium), W (Tungsten), and N (Nitrogen). This implies that higher concentrations of these elements may lead to lower martensite start temperatures.

The relationships identified here can be utilized to build predictive models for Ms based on the chemical composition of alloys.

#### Boxplot for Ms vs Carbon (C)

This boxplot graph illustrates the relationship between the Martensite Start Temperature (Ms) and Carbon content in percentage (%)

```{r}
ggplot(Martensite, aes(x = as.factor(C), y = Ms)) +
  geom_boxplot(fill = "red") +
  labs(title = "Boxplot of Martensite Ms vs Carbon", x = "Carbon (%)", y = "Martensite Start Temperature (Ms)")

```

#### Insights from the Boxplot

X-axis (Carbon %): This represents the percentage of carbon content in the material. The values are spread out across a wide range, indicating different carbon concentrations.

Y-axis (Martensite Start Temperature, Ms): This represents the temperature at which martensite begins to form during the cooling process. The temperature is measured in degrees Celsius.

As the carbon content increases, the Ms temperature decreases. This is typical, as higher carbon content usually lowers the Ms temperature in steels and alloys.


